---
title: "Data Engineering with Arrow Exercises"
execute:
  echo: true
  messages: false
  warning: false
---

# Schemas

```{r}
#| label: load-packages
library(arrow)
library(dplyr)
library(tictoc)
```

```{r}
#| label: open-dataset-seattle-csv
seattle_csv <- open_dataset(
  sources = here::here("data/seattle-library-checkouts.csv"),
  format = "csv"
)
```

::: {#exercise-schema .callout-tip}
# Data Types & Controlling the Schema

::: panel-tabset
## Problem

1.  The first few thousand rows of `ISBN` are blank in the Seattle Checkouts CSV file. Read in the Seattle Checkouts CSV file with `open_dataset()` and ensure the correct data type for `ISBN` is `<string>` instead of the `<null>` interpreted by Arrow.

2.  Once you have a `Dataset` object with the metadata you are after, count the number of `Checkouts` by `CheckoutYear` and arrange the result by `CheckoutYear`.

## Solution 1

```{r}
#| label: seattle-csv-schema-1
#| cache: true
seattle_csv <- open_dataset(
  sources = here::here("data/seattle-library-checkouts.csv"),
  format = "csv",
  skip = 1,
  schema(
    UsageClass = utf8(),
    CheckoutType = utf8(),
    MaterialType = utf8(),
    CheckoutYear = int64(),
    CheckoutMonth = int64(),
    Checkouts = int64(),
    Title = utf8(),
    ISBN = string(),
    Creator = utf8(),
    Subjects = utf8(),
    Publisher = utf8(),
    PublicationYear = utf8()
  )
)
```

or

```{r}
#| label: seattle-csv-schema-2
seattle_csv <- open_dataset(
  sources = here::here("data/seattle-library-checkouts.csv"),
  format = "csv",
  skip = 1,
  schema(
    UsageClass = utf8(),
    CheckoutType = utf8(),
    MaterialType = utf8(),
    CheckoutYear = int64(),
    CheckoutMonth = int64(),
    Checkouts = int64(),
    Title = utf8(),
    ISBN = utf8(),
    Creator = utf8(),
    Subjects = utf8(),
    Publisher = utf8(),
    PublicationYear = utf8()
  )
)
```

## Solution 2

The number of `Checkouts` by `CheckoutYear` arranged by `CheckoutYear`:

```{r}
#| label: seattle-csv-dplyr-1
#| cache: true
tic()
seattle_csv |>
  group_by(CheckoutYear) |>
  summarise(sum(Checkouts)) |>
  arrange(CheckoutYear) |> 
  collect()
toc()
```

or

```{r}
#| label: seattle-csv-dplyr-2
#| cache: true
tic()
seattle_csv |> 
  count(CheckoutYear, wt = Checkouts) |> 
  arrange(CheckoutYear) |> 
  collect()
toc()
```

Querying 42 million rows of data stored in a CSV on disk in \~10 seconds, not too bad.
:::
:::

# Parquet

```{r}
#| label: write-dataset-seattle-parquet
#| eval: false
seattle_parquet <- here::here("data/seattle-library-checkouts-parquet")

seattle_csv |>
  write_dataset(path = seattle_parquet,
                format = "parquet")
```

::: {#exercise-dataset .callout-tip}
# Parquet

::: panel-tabset
## Problems

1.  Re-run the query counting the number of `Checkouts` by `CheckoutYear` and arranging the result by `CheckoutYear`, this time using the Seattle Checkout data saved to disk as a single, Parquet file. Did you notice a difference in compute time?

## Solution 1

```{r}
#| label: seattle-parquet-dplyr-1
seattle_parquet <- here::here("data/seattle-library-checkouts-parquet")

tic()
open_dataset(seattle_parquet, 
             format = "parquet") |>
  group_by(CheckoutYear) |>
  summarise(sum(Checkouts)) |>
  arrange(CheckoutYear) |> 
  collect()
toc()
```

or

```{r}
#| label: seattle-parquet-dplyr-2
seattle_parquet <- here::here("data/seattle-library-checkouts-parquet")

tic()
open_dataset(seattle_parquet, 
             format = "parquet") |>
  count(CheckoutYear, wt = Checkouts) |> 
  arrange(CheckoutYear) |> 
  collect()
toc()
```

A *much* faster compute time for the query when the on-disk data is stored in the Parquet format.
:::
:::

# Partitioning

```{r}
#| label: write-dataset-seattle-partitioned
#| eval: false
seattle_parquet_part <- here::here("data/seattle-library-checkouts")

seattle_csv |>
  group_by(CheckoutYear) |>
  write_dataset(path = seattle_parquet_part,
                format = "parquet")
```

::: {#exercise-dataset .callout-tip}
# Partitioning

::: panel-tabset
## Problems

1.  Let's write the Seattle Checkout CSV data to a multi-file dataset just one more time! This time, write the data partitioned by `CheckoutType` as Parquet files.

2.  Now compare the compute time between our Parquet data partitioned by `CheckoutYear` and our Parquet data partitioned by `CheckoutType` with a query of the total number of checkouts in September of 2019. Did you find a difference in compute time?

## Solution 1

Writing the data:

```{r}
#| label: write-dataset-seattle-checkouttype
#| eval: false
seattle_checkouttype <- here::here("data/seattle-library-checkouts-type")

seattle_csv |>
  group_by(CheckoutType) |>
  write_dataset(path = seattle_checkouttype,
                format = "parquet")
```

## Solution 2

Total number of Checkouts in September of 2019 using partitioned Parquet data by `CheckoutType`:

```{r}
#| label: seattle-partitioned-other-dplyr
tic()
open_dataset(here::here("data/seattle-library-checkouts-type")) |> 
  filter(CheckoutYear == 2019, CheckoutMonth == 9) |> 
  group_by(CheckoutYear) |> 
  summarise(TotalCheckouts = sum(Checkouts)) |>
  collect() 
toc()
```

Total number of Checkouts in September of 2019 using partitioned Parquet data by `CheckoutYear` and `CheckoutMonth`:

```{r}
#| label: seattle-partitioned-partitioned-filter-dplyr
tic()
open_dataset(here::here("data/seattle-library-checkouts")) |> 
  filter(CheckoutYear == 2019, CheckoutMonth == 9) |> 
  group_by(CheckoutYear) |> 
  summarise(TotalCheckouts = sum(Checkouts)) |>
  collect() 
toc()
```

Faster compute time because the `filter()` call is based on the partitions.
:::
:::
