---
footer: "[🔗 posit.io/arrow](https://posit-conf-2023.github.io/arrow)"
logo: "images/logo.png"
execute:
  echo: true
format:
  revealjs: 
    theme: default
engine: knitr
---

# Data Manipulation---Part 2 {#data-manip-2}

```{r}
#| label: setup-data-manip-2
#| message: false
#| echo: false

library(arrow)
library(dplyr)
library(duckdb)
library(lubridate)
library(stringr)
library(tictoc)

nyc_taxi <- open_dataset(here::here("data/nyc-taxi/"))
```

# Joins

## Joining a reference table

```{r}
#| label: ref

vendors <- tibble::tibble(
  code = c("VTS", "CMT", "DDS"),
  full_name = c(
    "Verifone Transportation Systems",
    "Creative Mobile Technologies",
    "Digital Dispatch Systems"
  )
)

nyc_taxi %>%
  left_join(vendors, by = c("vendor_name" = "code")) %>%
  select(vendor_name, full_name, pickup_datetime) %>%
  head(3) %>%
  collect()
```

## Traps for the unwary

```{r}
#| label: lookup

nyc_taxi_zones <-
  read_csv_arrow(here::here("data/taxi_zone_lookup.csv")) %>%
  select(location_id = LocationID,
         borough = Borough)

nyc_taxi_zones
```

## Why didn't this work?

```{r}
#| label: join-fail
#| error: true

nyc_taxi |>
  left_join(nyc_taxi_zones, by = c("pickup_location_id" = "location_id")) |>
  collect()
```

## Schema for the `nyc_taxi` table

```{r}
#| label: get-schema

nyc_taxi$schema
```

## Schema for the `nyc_taxi_zones` table

```{r}
#| label: schema-2

arrow_table(nyc_taxi_zones)$schema
```

-   `pickup_location_id` is int64 in the `nyc_taxi` table
-   `location_id` is int32 in the `nyc_taxi_zones` table

## Take control of the schema

```{r}
#| label: zones-schema

nyc_taxi_zones_arrow <- arrow_table(
  nyc_taxi_zones, 
  schema = schema(location_id = int64(), borough = utf8())
)
```

-   `schema()` takes variable name / types as input
-   {arrow} has various "type" functions: `int64()`, `utf8()`, `boolean()`, `date32()` etc

## Take control of the schema

```{r}
#| label: zones-schema-2

nyc_taxi_zones_arrow <- arrow_table(
  nyc_taxi_zones, 
  schema = schema(location_id = int64(), borough = utf8())
)
nyc_taxi_zones_arrow$schema
```

## Prepare the auxiliary tables

```{r}
#| label: auxillary

pickup <- nyc_taxi_zones_arrow |>
  select(pickup_location_id = location_id,
         pickup_borough = borough)

dropoff <- nyc_taxi_zones_arrow |>
  select(dropoff_location_id = location_id,
         dropoff_borough = borough)
```

-   Join separately for the pickup and dropoff zones

```{=html}
<!-- 
    * simple example of a join
    * show a join with unmatched schemas
    * making the schemas match
-->
```
## Join and cross-tabulate

```{r}
#| label: join-crosstab

library(tictoc)

tic()
borough_counts <- nyc_taxi |> 
  left_join(pickup) |>
  left_join(dropoff) |>
  count(pickup_borough, dropoff_borough) |>
  arrange(desc(n)) |>
  collect()
toc()
```

<br>

2-3 minutes to join twice and cross-tabulate on non-partition variables, with 1.15 billion rows of data 🙂

## The results

```{r}
#| label: borough-counts

borough_counts
```

## Your Turn

1.  How many taxi pickups were recorded in 2019 from the three major airports covered by the NYC Taxis data set (JFK, LaGuardia, Newark)?

➡️ [Data Manipulation Part I Exercises Page](4_data_manipulation_2-exercises.html)

# Window functions

## What are window functions?

-   calculations within groups

## Grouped summaries

```{r}
#| label: grouped-summaries

fare_by_year <- nyc_taxi %>%
  filter(year %in% 2021:2022) %>%
  select(year, fare_amount)

fare_by_year %>%
  group_by(year) %>%
  summarise(mean_fare = mean(fare_amount)) %>% 
  collect()
```

## Window functions

```{r}
#| label: window-fail
#| error: true

fare_by_year %>%
  group_by(year) %>%
  mutate(mean_fare = mean(fare_amount)) %>% 
  collect()
```

## Window functions - via joins

```{r}
#| label: window-join
#| error: true

fare_by_year %>%
  left_join(
    nyc_taxi %>%
      group_by(year) %>%
      summarise(mean_fare = mean(fare_amount))
  ) %>% 
  arrange(desc(fare_amount)) %>%
  collect()
```

## Window functions - via duckdb

```{r}
#| label: window-duckdb
#| eval: false

fare_by_year %>%
  group_by(year) %>%
  to_duckdb() %>%
  mutate(mean_fare = mean(fare_amount)) %>% 
  to_arrow() %>%
  arrange(desc(fare_amount)) %>%
  collect()
```

## Your Turn

1.  How many trips in 2019 had a longer than average distance for that year?

➡️ [Data Manipulation Part I Exercises Page](4_data_manipulation_2-exercises.html)

## Custom functions

-   Not officially supported
-   Works for simple operations but not with bindings

## Custom functions - supported

```{r}
#| label: custom-1

millions <- function(x) x / 10^6

nyc_taxi |>
  group_by(vendor_name) %>%
  summarise(trips = n()) %>%
  mutate(
    trips_mil = millions(trips)
  ) %>%
  collect()
```

## Custom functions - not supported

```{r}
#| label: custom-fail
#| error: true

morning <- function(x) ifelse(lubridate::am(x), "morning", "afternoon")
nyc_taxi |>
  group_by(morning(pickup_datetime)) %>%
  count() %>%
  collect()
```

-   recommendation: write code as dplyr expressions instead of functions, or look up docs on user-defined functions for datasets (see `?register_scalar_function`)
