---
title: "Data Manipulation Part 2 - Exercises"
execute:
  echo: true
  messages: false
  warning: false
---

```{r}
#| label: load-packages
library(arrow)
library(dplyr)
library(duckdb)
```

```{r}
#| label: open-dataset
nyc_taxi <- open_dataset(here::here("data/nyc-taxi"))
```

::: {#exercise-udfs .callout-tip}
# User-defined functions

::: panel-tabset
## Problem

1.  Write a user-defined function which wraps the `stringr` function `str_replace_na()`, and use it to replace any `NA` values in the `vendor_name` column with the string "No vendor" instead. (Test it on the data from 2019 so you're not pulling everything into memory)

## Solution 1

```{r}
#| label: udf-preview

# Preview the distinct vendor names before we start
nyc_taxi |>
  filter(year == 2019) |> # smaller subset of the data
  distinct(vendor_name) |>
  collect()
```

```{r}
#| label: udf-solution
register_scalar_function(
  name = "replace_vendor_na",
  function(context, string) {
    stringr::str_replace_na(string, "No vendor")
  },
  in_type = schema(string = string()),
  out_type = string(),
  auto_convert = TRUE
)

vendor_names_fixed <- nyc_taxi |>
  mutate(vendor_name = replace_vendor_na(vendor_name)) 

# Preview the distinct vendor names to check it's worked
vendor_names_fixed |>
  filter(year == 2019) |> # smaller subset of the data
  distinct(vendor_name) |>
  collect()
```
:::
:::

::: {#exercise-joins .callout-tip}
# Joins

::: panel-tabset
## Problem

1.  How many taxi pickups were recorded in 2019 from the three major airports covered by the NYC Taxis data set (JFK, LaGuardia, Newark)? (Hint: you can use `stringr::str_detect()` to help you find pickup zones with the word "Airport" in them)

## Solution 1

```{r}
#| label: airport-pickup
pickup_location <- read_csv_arrow(here::here("data/taxi_zone_lookup.csv"))

pickup_location <- pickup_location |>
  select(
    pickup_location_id = LocationID,
    borough = Borough,
    pickup_zone = Zone
  ) 


pickup_location_arrow <- arrow_table(
  pickup_location, 
  schema = schema(
    pickup_location_id = int64(),
    borough = utf8(),
    pickup_zone = utf8()
  ))

nyc_taxi |>
  filter(year == 2019) |>
  left_join(pickup_location_arrow) |>
  filter(str_detect(pickup_zone, "Airport")) |>
  count(pickup_zone) |>
  collect()
```
:::
:::

::: {#exercise-window .callout-tip}
# Window functions

::: panel-tabset
## Problem

1.  How many trips in September 2019 had a longer than average distance for that month?

## Solution 1

### Option 1 - via DuckDB

```{r}
#| label: window-duckdb
nyc_taxi |>
  filter(year == 2019, month == 9) |>
  to_duckdb() |>
  mutate(mean_distance = mean(trip_distance)) |>
  to_arrow() |>
  filter(trip_distance < mean_distance) |>
  count() |>
  collect()
```

### Option 2 - via a join

```{r}
#| label: window-join
nyc_taxi |>
  filter(year == 2019, month == 9) |>
  left_join(
    nyc_taxi |>
      filter(year == 2019, month == 9) |>
      group_by(year) |>
      summarise(mean_distance = mean(trip_distance))
    ) |>
  filter(trip_distance < mean_distance) |>
  count() |>
  collect()
```
:::
:::
